### 2025-07-21 paper/sections/related.qmd
editorial pass until section 3.4

In a closely related work, @altmeyer2024faithful show that model averaging and, in particular, contrastive model objectives can produce more explainable and hence trustworthy models. The authors propose a way to generate counterfactuals that are maximally faithful to the model in that they are consistent with what the model has learned about the underlying data. Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by the model. Their counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them. The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your], a form of constrastive learning, tend to do well in this regard.

It helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are particularly effective when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the aforementioned risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations. Likewise, previous work of @schut2021generating found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have undergone adversarial training. The case for JEMs is even clearer: they optimize a hybrid objective that induces both high predictive performance and strong generative capacity [@grathwohl2020your], which resembles the idea of aligning models with plausible explanations and has inspired CT.

>>>>

In a closely related work, @altmeyer2024faithful show that model averaging and, in particular, contrastive model objectives can produce more explainable and hence trustworthy models. The authors propose a way to generate counterfactuals that are maximally faithful models in that they are consistent with what models have learned about the underlying data. Formally, they rely on tools from energy-based modelling [@teh2003energy] to minimize the contrastive divergence between the distribution of counterfactuals and the conditional posterior over inputs learned by a model. Their counterfactual explainer, *ECCCo*, yields plausible explanations if and only if the underlying model has learned representations that align with them. The authors find that both deep ensembles [@lakshminarayanan2016simple] and joint energy-based models (JEMs) [@grathwohl2020your], a form of constrastive learning, tend to do well in this regard.

It helps to look at these findings through the lens of representation learning with high degrees of freedom. Deep ensembles are approximate Bayesian model averages, which are particularly effective when models are underspecified by the available data [@wilson2020case]. Averaging across solutions mitigates the risk of overrelying on a single locally optimal representation that corresponds to semantically meaningless explanations. Likewise, previous work of @schut2021generating found that generating plausible ("interpretable") CEs is almost trivial for deep ensembles that have undergone adversarial training. The case for JEMs is even clearer: they optimize a hybrid objective that induces both high predictive performance and strong generative capacity [@grathwohl2020your], which resembles the idea of aligning models with plausible explanations and has inspired CT.
